Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-04-30 analysis/paper/paper.Rmd
add references and a sentence about social status
update the chronology para in excavation section and add two more references

The direct impacts of European colonialism on Indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local Indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in parts of Taiwan, the northeastern region was isolated from intensive direct contact by the Xueshan Mountains.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years [@Chen2007]. The sterility may be related to dry weather leading to site abandonment according to pollen analysis [@Lin2015]. These component divisions are based on the differences in the color and texture of the deposit. Whether these two components represent a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts relative to the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the supplementary information document.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, and are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Hsieh2009; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in shipping trade played an important role in introducing foreign jars to Taiwan. Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans that was widely adopted in European countries in the 16th century and spread to other regions through interactions [@Uccar2019]. We found that the presence of pipes in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). Those temporal diagnostic ceramics might have been continually used in later period that could mislead our chronology. Thus, we focus on identifying the certain context with high frequency of those ceramics that can reflect the early 17th century when they prevailed.

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrants in 1768 [@Chen1963; @Ke1993].

@@

The distribution and density of prestige goods across the residential area can provide information of social structure based on the assumption that distribution patterns observed from spatial data can reveal cultural processes by statistical methods [@Kintigh1982]. Point pattern analysis can assess whether the distribution of artifacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments in specific households that might result from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, because artifacts from Kiwulan lack exact piece-provenance data, we assigned each ornament to a random coordinate pair in the square it was recovered from. The next step was to divide the ornaments into three time periods. Finally we computed the density maps for each time period for comparison. Density values of artifacts per square meter were calculated for each cell. Here we use Kernel density estimations (KDE) for visualization and identification of spatial clusters [@Baxter1997], the core areas of ornaments and surrounding neighborhoods in this case. KDE is a approach of spatial analysis that computes the probability of the density of ornaments across space by creating a continuous, smooth density surface across space [@Bonnier2019; @Cortegoso2016].

>>>>

The direct impacts of European colonialism on Indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local Indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@Acabado2017; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in parts of Taiwan, the northeastern region was isolated from intensive direct contact by the Xueshan Mountains.

@@

The chronology of Kiwulan can be divided into two phases represented by a upper component (1350-1850 AD, 600-100 BP) and a lower component (650-1150 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years [@Chen2007]. The sterility may be related to dry weather leading to site abandonment according to pollen analysis [@Lin2015]. These component divisions are based on the differences in color and texture of the deposit and the content of artifacts. Whether these two components represent a culture continuity from the same people is still under debate [@Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. Based on these six units and copies of original excavation records, we have made some refinements of the original chronology to assign contexts relative to the pre-European, European, and the Chinese periods. The methods we used to devise the new chronology are described in the Supplementary Information document.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, and are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Dizon2016; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by Indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in shipping trade played an important role in introducing foreign jars to Taiwan.

Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made clay pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans. This custom was widely adopted in many European countries in the 16th century and spread to other regions through interactions [@Uccar2019]. We found that the presence of pipe bowls in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). It should be noted that the temporal indicators might have been introduced before the European arrival by Chinese traders that could mislead the interpretation of the start of European phase. However, the archaeological evidence shows that the layers with abundant trade ceramics match the 17th century according to excavation report and previous studies [@Hsieh2009; @Wang2011]. Thus, we focus on identifying the context with high frequency of those ceramics indicating the early 17th century when they prevailed.

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many styles and forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrants in 1768 [@Chen1963; @Ke1993]. The details about the distribution of the temporal indicators, accompanied with stratigraphy data, radiocarbon dates, and archaeological contexts for each sampled excavation unit are provided in Supplementary Information data.

@@

The distribution and density of prestige goods across the residential area can provide information of social structure based on the assumption that distribution patterns observed from spatial data can reveal cultural processes by statistical methods [@Kintigh1982]. The distribution pattern of artifacts can reveal social differences among the societies in many archaeological contexts [@Orser1988; @Pearson1993; @Wason2004]. Point pattern analysis can assess whether the distribution of artifacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments in specific households that might result from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, because artifacts from Kiwulan lack exact piece-provenance data, we assigned each ornament to a random coordinate pair in the square it was recovered from. The next step was to divide the ornaments into three time periods. Finally we computed the density maps for each time period for comparison. Density values of artifacts per square meter were calculated for each cell. Here we use Kernel density estimations (KDE) for visualization and identification of spatial clusters [@Baxter1997], the core areas of ornaments and surrounding neighborhoods in this case. KDE is a approach of spatial analysis that computes the probability of the density of ornaments across space by creating a continuous, smooth density surface across space [@Bonnier2019; @Cortegoso2016].
